NOPT048 Notes

### 2019-12-04 misc/study-notes/groups.tex
Added LING 360 and more group theory

Group theory is intimately connected to the study of symmetries of an object. Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself. This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $\Id_T$ and every bijection $f$ has an inverse bijection $f^{-1}$. In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections. For instance, $GL_n(\R) \subset \Sym(\R^n)$.

@@

Let $G$ be a group. We call a nonempty subset $H\subset G$ a {\it subgroup} and write $H\leq G$ provided that
\medskip
\item {a)} The set $H$ is closed under the multiplication operation of $G$.
\smallskip
\item {b)} Whenever $H$ contains an element $a\in G$, $H$ contains its inverse $a^{-1}$ as well.
\medskip
It is immediate from these requirements that a subgroup $H$ contains the identity element. Since $H$ is nonempty, it contains an element $h$ as well as its inverse $h^{-1}$. Then from closure of multiplication we conclude that $hh^{-1} = e\in H$.

@@

Now we must show that these $b\Z$ are all the possible subgroups. Let $H\leq \Z$. It is possible that $H$ contains only the identity $0$, in which case $H = 0\Z$. If not, let $b$ be the smallest positive integer contained in $H$. We know from closure that every multiple of $b$ is in $H$, so $b\Z\subset H$. Now let $h$ be any element in $H$. By the Euclidean division algorithm, we can divide $h$ by $b$ to get $h = mb + r$, where $mb$ is some multiple of $b$ and $r$ is a remainder lying in the range $0\leq r < b$. Since $r = (-mb) + h$, we have $r\in H$. But then necessarily we have $r=0$, since $b$ is the smallest positive integer in $H$. So $h$ is an integer multiple of $b$ and we have shown that $H \subset b\Z$. \slug

@@

If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called {\it cyclic} or {\it singly-generated}. More generally, if $S\subset G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
$$\langle S\rangle = \{s_1s_2\cdots s_n : n\in\N, s_i\in S\cup S^{-1}\}$$
is a subgroup of $G$, called the subgroup generated by $S$. If it so happens that $\langle S\rangle = G$, then we way that $S$ is a {\it generating set} of $G$.

@@

The following lemma describes the link between conjugacy and normal subgroups.

@@

On the other hand, suppose $H$ is a subgroup such that, for any $g\in G$ and $h\in H$, $ghg^{-1}\in H$. Then $gHg^{-1}\subset H$ and we can conclude that $gH\subset Hg$. But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subset H$ and $Hg \subset gH$. So $gH=Hg$ for all $g\in G$ and $H$ is normal.\slug

Let $G$ be a group, $H$ a subgroup of $G$, and let $S$ be a non-empty set of elements of $G$. The set of $h\in H$ such that $hSh^{-1} = H$ is called the {\it normaliser of $S$ in $H$}, denoted $N_H(S)$. The set of $h\in H$ such that $hsh^{-1} = s$ for all $s\in S$ is called the {\it centraliser of $S$ in $H$}, denoted $Z_H(S)$. It is easy to check that these are both subgroups of $H$. Note that if $S$ consists only of one element, then the centraliser and normaliser are equal and the inclusion $Z_H(S) \subset N_H(S)$ always holds. If $H = G$, then we may omit the subscript and simply call the subgroups the normaliser and centraliser of $S$.

@@

\section 2.6. The First Isomorphism Theorem

@@

\proof Let $h\in H$ and let $X$ denote $f^{-1}(h)$, the fibre of $h$. Now take $g\in G$ such that $f(g) = h$. Since $f(k) = e_H$ for all $k\in K$, we have $f(sk) = f(s)f(k) = h = f(k)f(s) = f(ks)$. So $sK\subset X$ and $Ks\subset H$. To prove the reverse inclusions, we let $x\in X$ (so $f(x) = h$). Then since $f$ is a homomorphism, we have $f(g^{-1}) = h^{-1}$ and
$$f(xg^{-1}) = f(x)f(g^{-1}) = hh^{-1} = e_H = h^{-1}h = f(g^{-1})f(x) = f(g^{-1}x).$$
This implies that $xg^{-1}\in K$, so multiplying on the right by $g$, we find that $x\in Kg$. Similarly, $g^{-1}x\in K$ so $x\in gK$.\slug

@@

\parenproclaim Theorem I (First isomorphism theorem). Let $f:G\rightarrow H$ be a surjective group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug

@@

@@

Consider a model example of a transitive action. Let $G$ be a group with a subgroup $H\leq G$; let $S$ be the set $\{aH : a\in G\}$ of left cosets of $H$. The group $G$ acts on $S$ by taking a coset $aH$ to $gaH$. This action is transitive: For any $a, a'$, there exists some $g$ such that $g(a) = a'$, so we can take any $aH$ to $a'H$ via $g$ as well. The closure property of subgroups gives us $G_H = H$, and the stabiliser of an arbitrary coset $aH$ is the set $\{g\in G : gaH = aH\}$. We find that this is exactly the conjugate subgroup $aHa^{-1}$. More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subset G$.

@@

>>>>

Group theory is intimately connected to the study of symmetries of an object. Let $T$ be a set and let $\Sym(T)$ denote the set of all bijections from $T$ to itself. This is called the {\it symmetric group} on $T$ as, under composition of functions, it obeys all the group axioms: It contains the identity transformation $\Id_T$ and every bijection $f$ has an inverse bijection $f^{-1}$. In some sense, the symmetric group is the most general group, because all other groups arise from adding restrictions to these bijections. For instance, $GL_n(\R) \subseteq \Sym(\R^n)$.

@@

Let $G$ be a group. We call a nonempty subset $H\subseteq G$ a {\it subgroup} and write $H\leq G$ provided that
\medskip
\item {a)} The set $H$ is closed under the multiplication operation of $G$.
\smallskip
\item {b)} Whenever $H$ contains an element $a\in G$, $H$ contains its inverse $a^{-1}$ as well.
\medskip
It is immediate from these requirements that a subgroup $H$ contains the identity element. Since $H$ is nonempty, it contains an element $h$ as well as its inverse $h^{-1}$. Then from closure of multiplication we conclude that $hh^{-1} = e\in H$.

@@

Now we must show that these $b\Z$ are all the possible subgroups. Let $H\leq \Z$. It is possible that $H$ contains only the identity $0$, in which case $H = 0\Z$. If not, let $b$ be the smallest positive integer contained in $H$. We know from closure that every multiple of $b$ is in $H$, so $b\Z\subseteq H$. Now let $h$ be any element in $H$. By the Euclidean division algorithm, we can divide $h$ by $b$ to get $h = mb + r$, where $mb$ is some multiple of $b$ and $r$ is a remainder lying in the range $0\leq r < b$. Since $r = (-mb) + h$, we have $r\in H$. But then necessarily we have $r=0$, since $b$ is the smallest positive integer in $H$. So $h$ is an integer multiple of $b$ and we have shown that $H \subseteq b\Z$. \slug

@@

If $G$ is a group containing an element $g$ such that $\langle g\rangle = G$, then $G$ is called {\it cyclic} or {\it singly-generated}. More generally, if $S\subseteq G$ and we let $S^{-1}$ denote the set $\{s^{-1} : s\in S\}$, then the set
$$\langle S\rangle = \{s_1s_2\cdots s_n : n\in\N, s_i\in S\cup S^{-1}\}$$
is a subgroup of $G$, called the subgroup generated by $S$. If it so happens that $\langle S\rangle = G$, then we way that $S$ is a {\it generating set} of $G$.

@@

Given a group $G$ and any set $S$ of elements from $G$, we can conjugate the entire set by an element:
$$gSg^{-1} = \{gsg^{-1} : s\in S$$
When the set $S$ is a subgroup, the conjugate set will also be a subgroup, but not necessarily the same subgroup $S$. It turns out that we can obtain an equivalent definition of normal subgroups in terms of conjugacy.

@@

On the other hand, suppose $H$ is a subgroup such that, for any $g\in G$ and $h\in H$, $ghg^{-1}\in H$. Then $gHg^{-1}\subseteq H$ and we can conclude that $gH\subseteq Hg$. But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subseteq H$ and $Hg \subseteq gH$. So $gH=Hg$ for all $g\in G$ and $H$ is normal.\slug

Let $G$ be a group, $H$ a subgroup of $G$, and let $S$ be a non-empty set of elements of $G$. The set of $h\in H$ such that $hSh^{-1} = H$ is called the {\it normaliser of $S$ in $H$}, denoted $N_H(S)$. The set of $h\in H$ such that $hsh^{-1} = s$ for all $s\in S$ is called the {\it centraliser of $S$ in $H$}, denoted $Z_H(S)$. It is easy to check that these are both subgroups of $H$. Note that if $S$ consists only of one element, then the centraliser and normaliser are equal and the inclusion $Z_H(S) \subseteq N_H(S)$ always holds. If $H = G$, we may omit the subscript and simply call the subgroups the normaliser and centraliser of $S$. It is clear that a subgroup $H$ of $G$ is normal if and only if $N(H) = G$.

\proclaim Theorem N. Let $S$ be a set of elements of a group $G$. If $H\leq G$, the index in $H$ of the normaliser of $S$ in $H$, $[H:N_H(S)]$, is exactly the number of conjugates of $S$ under $H$.

\proof Write $N = N_H(S)$ for short and suppose that $H$ is the disjoint union of cosets
$$H = N \cup h_1N \cup \cdots \cup h_kN,$$
where $k = [H:N_H(S)]$. Now suppose that $h_i$ and $h_j$ are elements of $H$ such that $h_iSh_i^{-1} = h_jSh_j^{-1}$. This is true if and only if $S = (h_j^{-1}h_i)S(h_j^{-1}h_i)^{-1}$. So $h_j^{-1}h_i$ is in $N_H(S) = N$ and $h_i\in h_jN$. So two conjugates of $S$ are the same if and only if the conjugating elements belong to the same left coset of $N$. Thus $[H:N_H(S)]$ is exactly the number of conjugates of $S$ under $H$.\slug

@@

\section 2.6. The Isomorphism Theorems

@@

\proof Let $h\in H$ and let $X$ denote $f^{-1}(h)$, the fibre of $h$. Now take $g\in G$ such that $f(g) = h$. Since $f(k) = e_H$ for all $k\in K$, we have $f(sk) = f(s)f(k) = h = f(k)f(s) = f(ks)$. So $sK\subseteq X$ and $Ks\subseteq H$. To prove the reverse inclusions, we let $x\in X$ (so $f(x) = h$). Then since $f$ is a homomorphism, we have $f(g^{-1}) = h^{-1}$ and
$$f(xg^{-1}) = f(x)f(g^{-1}) = hh^{-1} = e_H = h^{-1}h = f(g^{-1})f(x) = f(g^{-1}x).$$
This implies that $xg^{-1}\in K$, so multiplying on the right by $g$, we find that $x\in Kg$. Similarly, $g^{-1}x\in K$ so $x\in gK$.\slug

@@

\parenproclaim Theorem I (First Isomorphism Theorem). Let $f:G\rightarrow H$ be a surjective group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$ and there exists a group homomorphism $f':G/K \rightarrow H$.\slug

The second and third isomorphism theorems give further results on the relationships between specific types of quotient groups.

@@

\section ?.? Direct Products and Abelian Groups

[Under construction.]

@@

Consider a model example of a transitive action. Let $G$ be a group with a subgroup $H\leq G$; let $S$ be the set $\{aH : a\in G\}$ of left cosets of $H$. The group $G$ acts on $S$ by taking a coset $aH$ to $gaH$. This action is transitive: For any $a, a'$, there exists some $g$ such that $g(a) = a'$, so we can take any $aH$ to $a'H$ via $g$ as well. The closure property of subgroups gives us $G_H = H$, and the stabiliser of an arbitrary coset $aH$ is the set $\{g\in G : gaH = aH\}$. We find that this is exactly the conjugate subgroup $aHa^{-1}$. More generally, if $G$ acts transitively on a set $S$ and $g(s) = s'$, then $G_{s'} = gG_sg^{-1}\subseteq G$.

@@

Lagrange's Theorem, presented in Section 2.4, states that if $G$ is a group with $|G|=n$, then the order of any element of $G$ divides $n$. The converse of this theorem does not hold: There does not necessarily exist a subgroup with size $m$ for all $m$ dividing $n$. For example, $A_4$ has no subgroup of order 6. However, the results of this section provide a partial converse to Lagrange's Theorem. To this end, we first introduce the concept of double cosets.

Let $H$ and $K$ be subgroups, not necessarily distinct, of a group $G$. For a fixed element $g$ of $G$, the set
$$HgK = \{hgk : h\in H\ \hbox{and}\ k\in K\}$$
is called a {\it double coset}.

First we show that, just like ordinary cosets, double cosets partition a group. First, every element $g$ of a group belongs to some double coset of $H$ and $K$, namely $HgK$. Then we have the following lemma:

\proclaim Lemma D. Let $H$ and $K$ be subgroups of a group $G$. For elements $x,y\in G$, the cosets $HxK$ and $HyK$ are either equal or disjoint.

\proof If $HxK$ and $HyK$ are not disjoint, then we can find $g\in HxK \cap HyK$. Suppose that $g = h_1xk_1 = h_2yk_2$; so $x = h_1^{-1}h_2yk_2k_1^{-1}$ and $y = h_2^{-1}h_1xk_1k_2^{-1}$. Then for any $hxk\in HxK$,
$$hxk = hh_1^{-1}h_2yk_2k_1^{-1}k,$$
meaning that $HxK\subseteq HyK$. Likewise,
$$hyk = hh_2^{-1}h_1xk_1k_2^{-1}k$$
and $HyK\subseteq HxK$.\slug

A double coset $HgK$ contains left cosets of $K$ that are of the form $(hg)K$ as well as right cosets of $H$ of the form $H(gk)$. In fact, we can calculate the exact quantity of such left and right cosets.

\proclaim Lemma Q. The number of right cosets of $H$ in $HgK$ is $[gKg^{-1} : H\cap gKg^{-1}]$ and the number of left cosets of $K$ in $HgK$ is $[H : H\cap gKg^{-1}]$.

\proof We form a bijection $f:HgK\rightarrow HgKg^{-1}$ that takes an element $hgk$ to $hgkg^{-1}$. This creates a correspondence between left cosets $h(gKg^{-1})$ of $gKg^{-1}$ and left cosets $(hg)K$ of $K$ as well as a correspondence between right cosets $H(gkg^{-1})$ of $H$ and right cosets $H(gk)$ of $H$ in $HgK$.

Let $D = (H\cap gKg^{-1}) \leq gKg^{-1}$. Then $gKg^{-1}$ can be written as a union of disjoint right cosets
$$gKg^{-1} = D \cup Dx_2 \cup \cdots \cup Dx_n,$$
where $n = [gKg^{-1} : D]$. Each $x_i$ belongs to $gKg^{-1}$ and the claim is that $H, Hx_2, \ldots, Hx_n$ are exactly the right cosets of $H$ that are in $HgK$. These cosets are distinct since if $Hx_i = Hx_j$, then $x_ix_j^{-1}\in H$, but since both $x_i$ and $x_j$ are in $gKg^{-1}$, this means that $x_ix_j^{-1}\in D$ and $Dx_i = Dx_j$, contradicting our choice of coset representatives. Now every right coset of $H$ in $HgKg^{-1}$ is of the form $Hx$, where $x = dx_i$ for some $d\in D$. But since $D\subseteq H$, $Hx = Hdx_i = Hx_i$. So the number of right cosets of $H$ in $HgKg^{-1}$ is $[gKg^{-1} :  D] = [gKg^{-1} : H\cap gKg^{-1}]$ and by the bijection above, this is the number of right cosets of $H$ in $HgK$.

In a similar manner, it can be shown that the number of left cosets of $gKg^{-1}$ in $HgKg^{-1}$ is $[H:D] = [H:H\cap gKg^{-1}]$ and this is exactly the number of left cosets of $K$ in $HgK$.\slug

So much for double cosets. We saw earlier that if $n$ is the order of a group $G$ and $d$ divides $n$, there does not necessarily exist a subgroup of order $d$. However, if $d$ is prime or a power of a prime, then there is such a subgroup. We begin with a useful theorem, due to Cauchy.

\parenproclaim Theorem C (Cauchy). If $G$ is a group that has order dividing a prime $p$, then $G$ contains an element of order $p$.

\proof Let $G$ be a group with order $mp$, where $p$ is prime. The proof is by strong induction on $m$. If $m=1$, then $G$ is cyclic, generated by an element of order $p$.

Now suppose that the theorem holds for all smaller multiples of $p$. Then if $G$ contains a proper subgroup $H$ whose index $[G:H]$ is not divisible by $p$, then the order of $H$ is divisible by $P$ and, by induction, $H$ contains an element of order $p$. In the case that all proper subgroups of $G$ have indices divisible by $p$, we split $G$ into a union of disjoint conjugacy classes:
$$G = \{e\} \cup C_2 \cup \cdots \cup C_k$$
Suppose that each conjugacy class $C_i$ has order $n_i\geq 1$. Let $x_i$ be a representative from each $C_i$. Then $n = n_1 + \cdots + n_k$, where, by Theorem 1.4.N, $n_i = [G:Z(\{s_i\})]$. If $n_i \neq 1$, then $n_i$ is the index of a proper subgroup of $G$, and by hypothesis is divisible by $p$. Surely $n_1 = 1$, so the number of distinct $i$ for which $n_i = 1$ divides $p$. But $n_i = 1$ if and only if $s_i\in Z(G)$, so the centre of $G$ has order dividing $p$. By the Fundamental Theorem of Finite Abelian Groups, this implies that it contains an element of order $p$.\slug

Cauchy's Theorem tells us that if $p$ divides the order of $G$, at least one subgroup of $G$ has order $p$. The first of Sylow's theorems proves a stronger result.

[Sylow's Theorems.]
